Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-01-28 manuscript/manuscript.tex
Revised version :)

\par{     The authors are grateful to XXX for their helpful comments.  }
\par{Correspondence concerning this article should be addressed
to Amanda Chen, Email: zchengj@connect.ust.hk}
}

@@

Career trajectories for men and women often differ substantially, even
among those with similar qualifications. One source of such differences
may be the advice that people receive. Prior work has documented that
men receive more aspirational advice and women more risk-averse advice,
which may explain why women are less likely to apply for and ultimately
obtain more rewarding positions (Kanze et al., 2018). Research on gender
discrimination has focused on the role of unconscious bias (Greenwald et
al., 2009; Greenwald \& Banaji, 1995), but interventions that try to
reduce this bias have been largely unsuccessful (Chang et al., 2019;
Paluck \& Green, 2009). Recent work has proposed cognitive, rather than
unconscious, biases that can lead to gender discrimination through the
formation of false beliefs (Hagmann et al., 2024). Here, we propose a
novel source of differences in advice by drawing on recent work on
belief-based utility, expectation-based reference points (Köszegi, 2006;
Kőszegi \& Rabin, 2009), and the hedonic consequences of information
(Golman et al., 2017).

@@

People often go to great lengths to avoid unpleasant information,
deriving belief-based utility from maintaining favorable views of
themselves (Golman et al., 2017; Loewenstein \& Molnar, 2018;
\textbf{Ho2018?}). Moreover, they often prefer advisers who withhold bad
news from them (Shalvi et al., 2019) and punish those who communicate
bad news (John et al., 2019). Advisers, anticipating these preferences
and consequences face a dilemma: Should they provide the accurate but
potentially painful truth, or shade their recommendations to preserve
the advisee's mood and the adviser's own standing? Advisers who fear
they will be blamed or disliked for undermining someone's confidence may
reasonably avoid candid feedback, particularly if they don't incur a
cost when the advisee makes a mistake.

@@

We present the results of three preregistered experiments in which
participants are paired anonymously as advisers and advisees. Advisers
are incentivized to offer accurate advice, with bonus earnings depending
on the outcome of an advisee's decision. We study a setting in which
advisees can communicate their expectations to an adviser, and
experimentally manipulate whether advisers observe this expectation. Our
design simulates organizational contexts where mentors and managers
often have some sense of an employee's confidence or aspirations and
thus could adjust their feedback to avoid disappointing them.
Interpersonal considerations would likely be stronger in situations in
which the adviser and advisee have an existing relationship and when the
advice is delivered face-to-face. Moreover, advisers usually do not
suffer any direct costs when their flattering advice leads to a poor
outcome for the recipient, particularly when it is not clear what would
have happened under a counterfactual.\^{}{[}In some cases, however,
repeated interaction might also increase incentives for honest feedback.
Someone who is known to persistently give overoptimistic advice may be
viewed as less trustworthy in the long run.{]}

@@

We begin by examining the performance of the Stage 1 participants.
Because our default treatment takes place after participants completed
the ability quiz, we do not expect a difference in performance across
the Low and High Default treatments. Indeed, the two groups scored no
different from one another (5.1 and 5.18 for the Low Default and High
Default treatments, respectively, \(t(199) = 0.31\), \(p = .756\)). The
default manipulation affected their initial choices. In the ``Low
Default'' treatment, only 22\% of participants initially preferred
competing against the High Performer Group, compared to 65\% of
participants who saw the High Performer Group highlighted (\(\chi^2\)(1,
n = 201) = 36.62, p \textless{} .001).

@@

{\caption{{Advice to compete against the High Performer Group in Study
1. Displaying the non-binding choice of those for whom the Low Performer
Group was selected by default makes it more likely that they are advised
to compete against the High Performer Group (Column 1). Column 2
restricts the analysis to advisers who observed advisees' initial
non-binding choice and controls for advisees' score on the quiz.
Standard errors are clustered at the adviser
level.}{\label{tbl-study1regs}}}
\vspace{-20pt}}

\centering
\begin{talltblr}[         %% tabularray outer open
entry=none,label=none,
note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]},
column{2,3}={}{halign=c,},
column{1}={}{halign=l,},
hline{12}={1,2,3}{solid, black, 0.05em},
}                     %% tabularray inner close
\toprule
& (1) & (2) \\ \midrule %% TinyTableHeader
High Default               & \num{0.011}    & \num{0.025}*    \\
& (\num{0.013})  & (\num{0.011})   \\
Expectation Shown          & \num{-0.041}*  &                  \\
& (\num{0.016})  &                  \\
Expectation x High Default & \num{0.037}+   &                  \\
& (\num{0.019})  &                  \\
Performance                &                 & \num{0.168}***  \\
&                 & (\num{0.004})   \\
Constant                   & \num{0.389}*** & \num{-0.503}*** \\
& (\num{0.012})  & (\num{0.020})   \\
N                          & \num{10000}    & \num{5000}      \\
\bottomrule
\end{talltblr}

@@

We now examine whether advisers took the advisees' expectations into
account. Recall that advisees who were defaulted to competing against
the High Performer Group were indeed more likely to choose this group,
which, from the advisers' perspective, could signal high confidence in
their abilities. Our theory predicts that showing expectations should
lead to more flattering advice--- specifically, a recommendation to
compete against the High Performer Group--- for advisees with a high
default choice compared to those with a low default choice. Column 1 of
Table~\ref{tbl-study1regs} reports a linear probability model with
advice to compete against the high group as the outcome measure,
incorporating the default choice advisees saw, whether expectations were
shown to the adviser, and the interaction of the two. Because each
adviser made ten recommendations, we cluster standard errors at the
adviser level. As predicted, we find a marginally significant
interaction effect: advisees with a default choice to compete against
the High Performer Group (nudged to have high expectations, though not
necessarily holding them) were more likely to receive advice to compete
against stronger competitors when their expectations were shown compared
to when they were hidden. We reported a linear probability model on
advising to compete against the High Performer group for advisers in the
``Expectations'' treatment, controlling for the true performance of the
advisee in Column 2 of Table~\ref{tbl-study1regs}. We observe a
significant main effect of the default treatment, where a high default
increased the likelihood of receiving flattering advice. It is worth
noting that, since only 65 \% of advisees in the high default group
actually stuck with their default choice, this represents a conservative
test of the effect of high expectations on the likelihood of receiving
flattering advice.

@@

\caption{\label{fig-study1Advice}Advisees who were nudged to express
high expectations were more likely to receive advice to compete against
the High Performer Group.}

@@

When the quality of a decision does not depend on someone's
expectations, the adviser's awareness of the advisee's expectations
should not influence the advice given. However, in line with our
argument that advisers take into account the belief utility of the
advisee and prefer to flatter rather than disappoint, we find that
displaying the advisee's expectations does make a difference.
Specifically, advisees who expressed a preference for stronger
competitors were more likely to receive `flattering' advice, encouraging
them to pursue that option. One possible explanation for this finding is
that, rather than trying to flatter, advisers may simply be telling
advisees what they want to hear. In the next study, we aim to
investigate the mechanism behind this discrepancy in advice-giving when
expectations are revealed, exploring the question: Why do people give
flattering advice?

@@

So far, we have shown that displaying expectations influences the advice
given. Specifically, advisers were more likely to recommend competing
against a group of high performers to advisees with higher performance
expectations compared to those without. We now directly test whether
this effect is driven by interpersonal concerns. In particular, we
explore whether more flattering advice stems from a desire to be liked,
and whether advisees, in turn, like advisers more when they receive
flattering advice.

We conducted a three-stage experiment in which we experimentally
manipulated the incentives for advisers. We began by inviting a group of
advisees to upload photos of themselves (``selfies'') and informed them
that they would be rated on their attractiveness. We grouped them with
nine other participants of the same gender and recruited participants of
the opposing gender to rank them from most to least attractive and to
provide advice. Specifically, we asked them to advise the participant
they ranked as the 7th most attractive (4th least attractive) on what
rank they should bet they were ranked by a larger group of raters.
Advisers were randomly assigned to two treatments, receiving a bonus
payment either if the advisee guessed their rank accurately or if the
advisee evaluated the adviser as likeable as measured by an
unincentivized scale response. We hypothesize that advisers who want to
be liked by advisees will recommend that they bet on a lower rank,
i.e.~that they are more attractive.

@@

We recruited 300 participants from Prolific and, after asking
demographic questions, invited them to upload photos of themselves
(selfies) to be rated by other participants on attractiveness. 100 men
and 107 women agreed to do so and uploaded pictures that adhered to our
instructions (e.g., did not include other people). We selected the first
100 photos from women to arrive at a gender-balanced sample
(\(M_{\text{Age}}\) = 39.37 years). Participants were informed that
their selfies would be randomly grouped with those of nine other
participants of their gender and ranked in terms of attractiveness by a
group of new Prolific participants of the opposite gender. Lastly, they
guessed their rank (unincentivized).

Next, we recruited a new, gender-balanced sample for the role of
advisers. 472 participants from Prolific (\(M_{\text{Age}}\) =
41.0275424 years; 49.79\% Female) started by providing demographic
information, then were matched to a group of the opposite gender. They
then ranked participants from most to least attractive using a drop-down
menu next to each picture. Because of a limitation with the survey
software, participants were able to select the same rank for multiple
participants, rather than rank them uniquely. We remove 115 participants
who did not follow instructions and provide a unique ranking.

After submitting their ratings, they saw the photo of the participant
they had ranked as the 7th most attractive (i.e., the 4th least
attractive). They were reminded of the rank they had given to that
person and informed that this participant would be invited back and
could earn a \$1 bonus if they guessed their rank correctly. The rank
was determined by the aggregate ratings of all participants who had
ranked this group. Their task was to give advice to this participant
about the rank they should bet on based on having observed all ten
selfies and their own assessment. We randomly assigned participants to
one of two incentivization schemes. In the ``Accuracy'' treatment, they
received a bonus identical to the advisee: \$1 if they guessed their
rank correctly. In the ``Likeability'' treatment, we informed them that
they would be rated by the advisee on a 5-point Likert scale for how
likeable they thought they were. Each point on the scale would translate
to a bonus of 20 cents. They then selected a rank that they would
recommend the advisee to bet on.

Finally, we invited participants from Stage 1 and were ranked as the 7th
attractive by at least one adviser (so that they received advice) back
for the follow-up survey. Following our preregistration, we kept the
survey open for 7 days. In total, 146 participants (77 men, 69 women)
returned. They were reminded of the selfie they uploaded in Stage 1 and
informed that a group of 10 selfies, including theirs, had been rated by
other participants from Prolific. They then saw the advice from a
randomly selected adviser and made their estimate with a \$1 incentive
for guessing accurately. They then saw the rank they had been advised to
bet on one more time and were asked to rate the adviser's likability,
warmth, friendliness, good-naturedness, trustworthiness, and sincerity
on 5-point Likert scales (adapted from Fiske et al. (2007)).

@@

\caption{\label{fig-study2advice}Advisers were asked to suggest what
rank to bet on for someone they ranked as the 7th most attractive out of
ten people. When incentivized for accuracy, most of them advised betting
on 7, which was based on their evaluation of the recipient's
attractiveness. However, when incentivized for likability, many
participants gave flattering advice by suggesting betting on higher
ranks that imply a higher evaluation of attractiveness.}

@@

We begin by examining the prior beliefs of advisees who uploaded their
selfies. On average, men guessed they ranked 5.76 in their group of 10
and women guessed that they ranked 6.37. Participants overall
underestimate their attractiveness relative to the benchmark average of
5.5. Moreover, women do so more than men \(t(205) = 2.15\),
\(p = .033\). Notably, people's self-perceptions correlated strongly
with the aggregate ratings of the advisers (0.43, \(t(193) = 6.70\),
\(p < .001\)). However, there was substantial heterogeneity in
perceptions of attractiveness. Of the 200 participants, 149 were ranked
as 7th most attractive by at least one adviser. On average, men in this
subset estimated they were ranked 5.94th and women estimated they were
ranked 6.59th (\(t(143) = 1.94\), \(p = .054\)).

Next, we turn our attention to the advisers
(Figure~\ref{fig-study2advice}). In the Accuracy condition, those
uploading selfies were advised to bet on rank 6.19. Notably, this is
significantly more attractive than the 7th rank those advisers had
themselves guessed just on the prior screen (\(t(234) = -8.82\),
\(p < .001\)). This suggests that even when incentivized for accuracy,
participants offered flattering advice.\footnote{This shading could be
  due to concerns of avoiding disappointment. However, it could also be
  that advisers are uncertain about the rankings they had given and make
  a recommendation that combines their own belief with a uniform prior.}
Importantly, and as predicted, we find that advisers in the Likeable
treatment recommend betting on a lower rank, communicating that they
think the participant in the selfie is more attractive (5.38,
\(t(470) = 5.44\), \(p < .001\)). This suggests that advisers inferred
that advising someone that they were more attractive would make the
adviser appear more likeable and therefore provided advice that
communicated a more favorable impression of the participants'
attractiveness. The distribution shown in Figure~\ref{fig-study2advice}
shows that participants do not simply tell participants that they are
the most attractive person in the group. They may infer that flattering
advise needs to be somewhat realistic to be believable. We return to
this in the general discussion.

@@

{\caption{{When individuals receive advice that implies a high level of
attractiveness (lower rank), they tend to perceive the advice giver as
more likable (Column 1), warm (Column 2). Column 3 shows that advisors
are rated as more trustworthy when they advice lower ranks, but this
relationship is only directional.}{\label{tbl-study2regs}}}
\vspace{-20pt}}

@@

\caption{\label{fig-study2ratingByGroup}After participants received
advice that suggested betting on a rank that was lower than, equal to,
or higher than their self-expectations, they evaluated advisors based on
their likability and warmth. Participants rated advisors who gave
flattering advice and delivered a high evaluation of their
attractiveness as more likable and warm than those whose advice implied
a lower evaluation of their attractiveness.}

@@

As seen in Figure~\ref{fig-study2rating}, we found that advisees who
suggested that the advisee was more attractive (lower rank) were indeed
rated as more likeable and warm (b = -0.109, p \textless{} .05; b =
-0.133, p \textless{} .001, respectively; See Columns 1 and 2 of
Table~\ref{tbl-study2regs}). When using their self-expectations as a
reference point, advisors who suggest a rank lower than that point are
perceived as less likable and warm (see
Figure~\ref{fig-study2ratingByGroup}). Interestingly, these benefits are
not at the cost of being hypocritical; advisers who recommend a more
favorable rank are viewed as no less trustworthy (Column 3 of
Table~\ref{tbl-study2regs}).

We were not powered to do a comparison across the two experimental
groups and did not preregister such a difference. Indeed, we find no
difference in likeability and warmth across the two treatments
(\(t(144) = 1.09\), \(p = .275\), and \(t(144) = 0.58\), \(p = .562\),
respectively). We also assess the quality of advice by measuring the
discrepancy between the advised rank and the final rank. Our analysis
revealed no significant difference between the two treatments in terms
of how helpful the advice was for advisees in making accurate guesses,
\(t(144) = -0.29\), \(p = .776\).

@@

We begin by first recruiting a sample of 50 participants from Prolific
to complete a 10-question multiple choice mathematics quiz. The
questions were taken from a paper-version of the ASVAB standardized
exam, such that answers were not available online. Participants had five
minutes to answer the quiz and were paid 10 cents for each correctly
answered question. Like before, we define top 20 scores as ``High
Performers'' and the bottom 20 scorers as the ``Low Performers.'' On
average, participants answered 4.42 questions correctly, High Performers
scored between 5 and 10, and Low Performers scored between 0 and 3. In
order to anchor the expectations of participants in our main experiment,
we simulated 1,000 pairings of groups of five participants, with the 5th
percentile of groups scoring an average of 2.6 and the 95 th percentile
scoring an average of 6.4. We report these averages to participants in
the Low Expectations and High Expectations treatment, respectively.

We then recruited 1,002 participants for the role of advisees in Stage 1
of our main experiment. To arrive at a gender-balanced sample, we
dropped the last two male participants to complete the survey, ending up
with a sample of 500 men and 500 women (\(M_{\text{Age}}\) = 42.16).
Participants completed the same 10-item mathematics quiz as the earlier
participants and were informed that their performance would affect their
bonus earnings in a follow-up stage to be conducted a few days later.
After completing the quiz, we informed them of the average score of a
group of five participants from the preliminary survey. We randomly
assigned them to learn about the 5th percentile of groups, which scored
2.6. (``Low Expectations'' treatment) or the 95th percentile of groups,
which scored 6.4 (``High Expectations'' treatment). Participants then
made a guess (unincentivized) about how many questions they think they
answered correctly. The survey concluded with basic demographic
questions.

We then recruited 1,000 participants for Stage 2, placing them in the
role of advisers. We began by informing them of the mathematics quiz
that participants in the preliminary study and Stage 1 had completed,
and informed them of the average score of all participants in the
preliminary study. Advisers had to recommend whether an advisee should
compete against the Low Performers or the High Performers (we used these
terms in the survey). We anticipated that being told to compete against
High Performers is more flattering and hence being told to compete
against the Low Performers would be disappointing if one had expected to
do well. Advisees could earn a bonus if their score was equal to or
higher than that of a randomly selected member from their chosen group.
Competing against the High Performers was more challenging, resulting in
an average bonus of 50 cents, compared to 30 cents when competing
against other groups.

Like before, advisers were randomly assigned to one of two treatments.
In the ``Baseline'' treatment, they only observed the score of the
advisee on the mathematics quiz. In the ``Expectation'' treatment, they
observed the score as well as the advisee's guess for how many questions
they answered correctly. We want to emphasize that sice the outcome is
determined only by the past score on the quiz, the advisee's guess is
immaterial to which group they should compete against. Moreover, in
neither treatment did they receive any demographic information about
their advisee. Participants gave recommendations to 10 advisees, which
unbeknownst to them were five men and five women matched to have
identical performance on the test.\footnote{We made this decision to
  account for the possibility of gender differences in performance.}
They were informed that if their advice was shown to a participant who
returned for the follow-up survey, they would receive the identical
bonus as that participant. The survey then concluded with basic
demographic questions.

>>>>

\par{       }
\par{Correspondence concerning this article should be addressed
to Amanda Chen, Email: zchengj@connect.ust.hk}
}

@@

Career trajectories for men and women often differ substantially, even
among those with similar qualifications. One source of such differences
may be the advice that people receive. Prior work has documented that
men receive more aspirational advice and women more risk-averse advice,
which may explain why women are less likely to apply for and ultimately
obtain more rewarding positions (Kanze et al., 2018). Research on gender
discrimination has focused on the role of unconscious bias (Banaji \&
Greenwald, 1995; Nosek et al., 2009), but interventions that try to
reduce this bias have been largely unsuccessful (Chang et al., 2019;
Paluck \& Green, 2009). Recent work has proposed cognitive, rather than
unconscious, biases that can lead to gender discrimination through the
formation of false beliefs (Hagmann et al., 2024). Here, we propose a
novel source of differences in advice by drawing on recent work on
belief-based utility, expectation-based reference points (Kőszegi \&
Rabin, 2006, 2009), and the hedonic consequences of information (Golman
et al., 2017).

@@

People often go to great lengths to avoid unpleasant information,
deriving belief-based utility from maintaining favorable views of
themselves (Golman et al., 2017; Ho et al., 2021; Loewenstein \& Molnar,
2018). Moreover, they often prefer advisers who withhold bad news from
them (Shalvi et al., 2019) and punish those who communicate bad news
(John et al., 2019). Advisers, anticipating these preferences and
consequences face a dilemma: Should they provide the accurate but
potentially painful truth, or shade their recommendations to preserve
the advisee's mood and the adviser's own standing? Advisers who fear
they will be blamed or disliked for undermining someone's confidence may
reasonably avoid candid feedback, particularly if they don't incur a
cost when the advisee makes a mistake.

@@

We present the results of three preregistered experiments in which
participants are paired anonymously as advisers and advisees. Advisers
are incentivized to offer accurate advice, with bonus earnings depending
on the outcome of an advisee's decision. We study a setting in which
advisees can communicate their expectations to an adviser, and
experimentally manipulate whether advisers observe this expectation. Our
design simulates organizational contexts where mentors and managers
often have some sense of an employee's confidence or aspirations and
thus could adjust their feedback to avoid disappointing them.
Interpersonal considerations would likely be stronger in situations in
which the adviser and advisee have an existing relationship and when the
advice is delivered face-to-face. Moreover, advisers usually do not
suffer any direct costs when their flattering advice leads to a poor
outcome for the recipient, particularly when it is not clear what would
have happened under a counterfactual.\footnote{In some cases, however,
  repeated interaction might also increase incentives for honest
  feedback. Someone who is known to persistently give overoptimistic
  advice may be viewed as less trustworthy in the long run.}

@@

We begin by examining the performance of the Stage 1 participants.
Because our default treatment takes place after participants completed
the ability quiz, we do not expect a difference in performance across
the Low and High Default treatments. Indeed, the two groups score no
different from one another (5.10 and 5.18 for the Low Default and High
Default treatments, respectively, \(t(199) = 0.31\), \(p = .756\)). As
expected, the default manipulation changes their initial choices. In the
``Low Default'' treatment, only 22\% of participants prefer to compete
against the High Performer Group, compared to 65\% of participants in
the ``High Default'' treatment (\(\chi^2\)(1, n = 201) = 36.62, p
\textless{} .001).

\linespread{1}

@@

{\caption{{Advice to compete against the High Performer Group in Study
1. Displaying the non-binding choice of those for whom the Low Performer
Group was selected by default makes it less likely that they are advised
to compete against the High Performer Group (Column 1). Column 2
restricts the analysis to advisers who observe advisees' initial
non-binding choice and controls for advisees' score on the quiz.
Standard errors are clustered at the adviser
level.}{\label{tbl-study1regs}}}\vspace{0pt}
}

\centering
\begin{talltblr}[         %% tabularray outer open
entry=none,label=none,
note{}={+ p \num{< 0.1}, * p \num{< 0.05}, ** p \num{< 0.01}, *** p \num{< 0.001}},
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]},
column{3}={}{halign=c,},
column{1}={}{halign=l,},
cell{2}{2}={}{halign=c,},
cell{3}{2}={}{halign=c,},
cell{4}{2}={}{halign=c,},
cell{5}{2}={}{halign=c,},
cell{6}{2}={}{halign=c,},
cell{7}{2}={}{halign=c,},
cell{8}{2}={}{halign=c,},
cell{9}{2}={}{halign=c,},
cell{10}{2}={}{halign=c,},
cell{11}{2}={}{halign=c,},
cell{12}{2}={}{halign=c,},
cell{13}{2}={}{halign=c,},
cell{1}{2}={c=2,}{halign=c, halign=c,},
hline{13}={1,2,3}{solid, black, 0.05em},
}                     %% tabularray inner close
\toprule
& Advice to Compete Against High Performers &  \\ \cmidrule[lr]{2-3}
& All Advice & Expectations Shown \\ \midrule %% TinyTableHeader
High Default               & \num{0.011}    & \num{0.025}*    \\
& (\num{0.013})  & (\num{0.011})   \\
Expectation Shown          & \num{-0.041}*  &                  \\
& (\num{0.016})  &                  \\
Expectation x High Default & \num{0.037}+   &                  \\
& (\num{0.019})  &                  \\
Score                      &                 & \num{0.168}***  \\
&                 & (\num{0.004})   \\
Constant                   & \num{0.389}*** & \num{-0.503}*** \\
& (\num{0.012})  & (\num{0.020})   \\
N                          & \num{10000}    & \num{5000}      \\
\bottomrule
\end{talltblr}

@@

We now examine whether advisers take the advisees' expectations into
account. Advisers observe the true performance of the advisee, such that
expectations about performance add no additional information. However,
because participants who score higher are also more likely to pick the
High Performer Group, we use our Default treatment as an instrument that
reflect higher expectations. We show linear probability models for the
decision of the Advisers, which is equal to 1 if they recommend
competing against the High Performer Group and 0 if they recommend the
Low Performer Group. In Column 1 of Table~\ref{tbl-study1regs}, we show
our main specificiation with the experimental assignment for Advisees
(High Performer Default vs.~Low Performer Default) and Advisers
(Expectation Shown vs.~Expectation Hidden), as well as their
interaction. Because each adviser made ten recommendations, we cluster
standard errors at the adviser level. We find that displaying
expectations of those in the Low Performer Default reduces
recommendations to compete against the Top group by 4 percentage points,
or about 10\% relative to when expectations are hidden. This decrease is
marginally smaller for advisees with a default choice to compete against
the High Performer Group (nudged to have high expectations, though not
necessarily holding them), and whose advice did not differ based on
whether their expectations were shown or hidden. We reported a linear
probability model on advising to compete against the High Performer
group for advisers in the ``Expectations'' treatment, controlling for
the true performance of the advisee in Column 2 of
Table~\ref{tbl-study1regs}. We observe a significant main effect of the
default treatment, where a high default increased the likelihood of
receiving flattering advice. It is worth noting that, since only 65 \%
of advisees in the high default group actually stuck with their default
choice, this represents a conservative test of the effect of high
expectations on the likelihood of receiving flattering advice.

@@

\caption{\label{fig-study1Advice}Advice given in Study 1. Advisees who
were nudged to express high expectations were more likely to receive
advice to compete against the High Performer Group.}

@@

When an adviser knows an outcome that an advisee can only estimate,
learning this estimate should not change the advice that is given.
However, in line with our argument that advisers take into account the
belief utility of the advisee and prefer to avoid disappointing them, we
find that providing information about expectations makes a difference.
Specifically, advisees who expressed a preference for stronger
competitors were more likely to receive flattering advice, encouraging
them to pursue that option. This occurs even though advisers are
incentivized to give good advice.

@@

We next examine if advisers believe that flattering advice will improve
how they are perceived, and if so, whether this perception is true. We
do this by manipulating the incentives for advisers, who receive a bonus
either based on how they are evaluated by an advisee, or a bonus for the
advisee's accuracy.

The study takes place in three stages. First, a group of advisees upload
photos of themselves (``selfies'') and are grouped with nine other
participants of the same sex We then recruit participants of the
opposing sex to rank them from most to least attractive and to provide
advice. Specifically, they advise the participant they ranked as the 7th
most attractive (i.e., 4th least attractive) on what rank they should
bet they were ranked by a larger group of raters. Advisers were randomly
assigned to two treatments, receiving a bonus payment either if the
advisee guessed their rank accurately or if the advisee evaluated the
adviser as likeable as measured by a scale response. We hypothesize that
advisers incentivized to be liked will recommend betting on a lower
rank, i.e.~that the advisee is more attractive.

@@

We recruit 300 participants from Prolific and, after asking demographic
questions, invite them to upload photos of themselves (selfies) to be
rated by other participants on attractiveness. We obtain selfies from
100 men and 107 women adhering to our instructions (e.g., did not
include other people). In line with our preregistration, we select the
first 100 photos from women to arrive at a gender-balanced sample
(\(M_{\text{Age}}\) = 39.37 years). Participants are informed that their
selfies will be randomly grouped with those of nine other participants
of their sex and ranked in terms of attractiveness by a group of new
Prolific participants of the opposite sex. They then make an
unincentivized guess of their rank.

Next, we recruit 472 participants from Prolific for the role of advisers
(\(M_{\text{Age}}\) = 41.0275424 years; 49.79\% Female). We first
collect demographic information, then match them to a group of the
opposite sex. They rank selfies from ten participants from most to least
attractive using a drop-down menu next to each picture. Because of a
limitation with the survey software, we could not validate that each
rank is given only once, and we remove 115 participants who failed to
follow instructions and did not provide a complete ranking.\footnote{Moreover,
  we could not collect data from any participant who did not select any
  participant as the 7th most attractive participant, as that image
  would be shown on subsequent pages.}

After submitting their ratings, participants see the photo of the
participant they ranked as the 7th most attractive (i.e., the 4th least
attractive). We remind them of the rank they have just given to that
person and inform them that this participant will be invited back and
can earn a \$1 bonus if they correctly guess their rank. The rank is
determined by the aggregate ratings of all participants who have ranked
this group. Because the participant does not observe the other nine
people in the group, they would depend on the adviser's recommendation
along with their own assessment. We randomly assign advisers to one of
two incentivization schemes. In the ``Accuracy'' treatment, they receive
a bonus identical to the advisee: \$1 if they guess their rank
correctly. In the ``Likeability'' treatment, we inform them that the
advisee will rate them on a 5-point Likert scale on how likeable they
thought they are. Each point on the scale would translate to a bonus of
20 cents. Advisers then select a rank (from 1 to 10) that they recommend
the advisee to bet on.

Finally, we invite participants from Stage 1 who were ranked as the 7th
most attractive participant by at least one adviser (so that they
received advice) back for the follow-up survey. Following our
preregistration, we keep the survey open for 7 days. In total, 146
participants (77 men, 69 women) complete the follow-up survey. We remind
them of the selfie they uploaded in Stage 1 and that a group of 10
selfies, including theirs, had been rated by other participants from
Prolific. They then see the advice from a randomly selected adviser and
make their own guess with a \$1 incentive for accuracy. Finally, they
rate the adviser's likability, warmth, friendliness, good-naturedness,
trustworthiness, and sincerity on 5-point Likert scales (adapted from
Fiske et al., 2007).

@@

\caption{\label{fig-study2advice}Participants in Study 2 give advice to
the person they rank 7th most attractive out of 10. When incentivized
for accuracy, the majority recommends betting on that rank. However,
when incentivized for likeability, participants inflate their advice to
flatter the recipient as more attractive.}

@@

We begin by examining the prior beliefs of advisees who uploaded their
selfies. On average, men guess they rank 5.76 in their group of 10 and
women guess that they rank 6.37 (\(t(205) = 2.15\), \(p = .033\)).
Notably, people's self-perceptions correlate strongly with the aggregate
ratings of the advisers (0.43, \(t(193) = 6.70\), \(p < .001\)).
However, there is substantial heterogeneity in perceptions of
attractiveness. Of the 200 participants, 149 are ranked as 7th most
attractive by at least one adviser. On average, men in this subset
estimate they rank 5.94th and women estimate they rank 6.59th
(\(t(143) = 1.94\), \(p = .054\)).

Next, we turn our attention to the advisers (see
Figure~\ref{fig-study2advice}). In the Accuracy condition, those
uploading selfies are on average advised to bet on rank 6.19. Notably,
this is significantly more attractive than the 7th rank those advisers
had themselves guessed just on the prior screen (\(t(234) = -8.82\),
\(p < .001\)). This suggests that even when incentivized for accuracy,
participants offer flattering advice.\footnote{This shading could be due
  to concerns of avoiding disappointment. However, it could also be that
  advisers are uncertain about the rankings they have given and make a
  recommendation that combines their own belief with a uniform prior.
  Therefore, our analyses focus on the difference between the two
  conditions.} Importantly, and as predicted, we find that advisers in
the Likeable treatment recommend betting on a lower rank, communicating
that they think the participant in the selfie is more attractive (5.38,
\(t(470) = 5.44\), \(p < .001\)). This suggests that advisers infer
flattering someone with pleasant advice would make the adviser appear
more likeable. The distribution shown in Figure~\ref{fig-study2advice}
shows that participants do not simply tell participants that they are
the most attractive person in the group. They may infer that flattering
advise needs to be somewhat realistic to be believable. We return to
this in the general discussion.

@@

{\caption{{When individuals receive advice that implies a high level of
attractiveness (lower rank), they tend to perceive the advice giver as
more likable (Column 1), warm (Column 2). Column 3 shows that advisors
are rated as more trustworthy when they advice lower ranks, but this
relationship is only directional.}{\label{tbl-study2regs}}}\vspace{0pt}
}

@@

\caption{\label{fig-study2ratingByGroup}After participants receive
advice that suggests betting on a rank that was lower than, equal to, or
higher than their own guess, they evaluate advisors based on their
likability and warmth. Participants rate advisors who give flattering
advice and deliver a high evaluation of their attractiveness as more
likable and warm than those whose advice implied a lower evaluation of
their attractiveness.}

@@

As seen in Figure~\ref{fig-study2rating}, we advisers who suggested that
the advisee was more attractive (lower rank) were indeed rated as more
likeable and warm (b = -0.109, p \textless{} .05; b = -0.133, p
\textless{} .001, respectively; See Columns 1 and 2 of
Table~\ref{tbl-study2regs}). When using their own initial guess as a
reference point, advisers who suggest a worse rank are perceived as less
likable and warm (see Figure~\ref{fig-study2ratingByGroup}).
Interestingly, these benefits are not at the cost of sincerity; advisers
who recommend a more favorable rank are viewed as no less trustworthy
(Column 3 of Table~\ref{tbl-study2regs}).

We are not powered to do a comparison across the two experimental groups
and did not preregister such a difference. Indeed, we find no difference
in likeability and warmth across the two treatments (\(t(144) = 1.09\),
\(p = .275\), and \(t(144) = 0.58\), \(p = .562\), respectively). We
also assess the quality of advice by measuring the discrepancy between
the advised rank and the correct rank. We find no difference in accuracy
across the advice treatments (\(t(144) = -0.29\), \(p = .776\)).

@@

We begin by first recruiting a sample of 50 participants from Prolific
to complete a 10-question multiple choice mathematics quiz. The
questions are taken from a paper-version of the ASVAB standardized exam,
such that answers are not available online. Participants have five
minutes to answer the quiz and are paid 10 cents for each correctly
answered question. Like in Study 1, we define the top 20 scores as
``High Performers'' and the bottom 20 scorers as the ``Low Performers.''
On average, participants answered 4.42 questions correctly, High
Performers scored between 5 and 10, and Low Performers scored between 0
and 3. In order to anchor the expectations of participants in our main
experiment, we simulated 1,000 pairings of groups of five participants,
with the 5th percentile of groups scoring an average of 2.6 and the 95
th percentile scoring an average of 6.4. We report these averages to
participants in the Low Expectations and High Expectations treatment,
respectively.

We then recruit 1,002 participants for the role of advisees in Stage 1
of our main experiment. To arrive at a gender-balanced sample, we drop
the last two male participants to complete the survey, ending up with a
sample of 500 men and 500 women (\(M_{\text{Age}}\) = 42.16).
Participants complete the same 10-item mathematics quiz as the earlier
participants and are informed that their performance would affect their
bonus earnings in a follow-up stage to be conducted a few days later.
After completing the quiz, we inform them of the average score of a
group of five participants from the preliminary survey. We randomly
assign them to learn about the 5th percentile of groups, which scored
2.6. (``Low Expectations'' treatment) or the 95th percentile of groups,
which scored 6.4 (``High Expectations'' treatment). Participants then
make an unincentivized guess about how many questions they think they
answered correctly. The survey concludes with basic demographic
questions.

We then recruit 1,000 participants for Stage 2, placing them in the role
of advisers. We begin by informing them of the mathematics quiz that
participants in the preliminary study and Stage 1 have completed, and
inform them of the average score of all participants in the preliminary
study. Advisers have to recommend whether an advisee should compete
against the Low Performers or the High Performers (we use these terms in
the survey). We anticipate that being told to compete against High
Performers is more flattering and hence being told to compete against
the Low Performers would be disappointing if one had expected to do
well. Advisees would earn a bonus if their score is equal to or higher
than that of a randomly selected member from their chosen group. They
earn 50 cents if they outperform a High Performer and 30 cents if they
outperform a Low Performer.

Advisers are randomly assigned to one of two treatments. In the
``Baseline'' treatment, they only observe the score of the advisee on
the mathematics quiz. In the ``Expectation'' treatment, they observe the
score as well as the advisee's guess for how many questions they
answered correctly. Since the outcome is determined only by the past
score on the quiz, the advisee's guess is not informative for the
recommendation, which can be based on the revealed score. Moreover, in
neither treatment did they receive any demographic information about the
advisees. Participants give recommendations to 10 advisees, which
unbeknownst to them are five men and five women matched to have
identical performance on the test.\footnote{We made this decision to
  account for the possibility of gender differences in performance.}
They are informed that if their advice is shown to a participant who
returns for the follow-up survey, they will receive the identical bonus
as that participant. The survey then concludes with basic demographic
questions.
